---
title:  "RCpp introduction"
subtitle: "FIMS Implementation Team Workshop"
author: "Christine Stawitz<br>NOAA Fisheries, OST<br>Seattle, WA<br> 2022-03-23"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "slides-style.css", "tufte.css"]
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---
layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-tile-view, echo=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()
```

---
# Sources
- "Advanced `R`"  by Hadley Wickham (http://adv-r.had.co.nz/Rcpp.html)
- "`Rcpp`: Seamless `R` and `C++` integration" by Dirk Eddelbutel and Romain Francois, Journal of Statistical Software (https://www.jstatsoft.org/article/view/v040i08)
- "`Rcpp` for everyone" by Masaki E. Tsuda (https://teuder.github.io/rcpp4everyone_en/)
- "Extending `Rcpp`" vignette by EddelButel and Francois

---

# What is `RCpp` and why should I care?

.bluebox[## Definition
"The `RCpp` package provides a consistent API for seamlessly accessing, extending or modifying R objects at the C++ level" - Eddelbuttel and Francois.
]

.yellowbox[## API
a set of functions and procedures allowing the creation of applications that access the features or data of an operating system, application, or other service.]

* One of the most widely used `R` extensions (over 1000 packages)
* With very minimal knowledge of `C++` you can speed up function runs A LOT
- The most efficient `R` functions are written in `C++` and called from `R`


---
# Writing `C++` functions in `R`

* You can write `C++` functions inline in your R code using `cppFunction`
* You can also compile single lines of code directly using `evalCpp`

```{r}
library(Rcpp)
cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')
# add works like a regular R function
add
add(1, 2, 3)
evalCpp("std::numeric_limits<double>::max()")
```

---
# Calling `C++` from `R`

.pull-left[
mean.cpp
```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;

  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}

```
]

.pull-right[
* `#include <Rcpp.h>` allows you to call `Rcpp` syntax in the file. You can omit `using namespace Rcpp` if you use `::` i.e. `Rcpp::NumericVector` 
* `// [[Rcpp::export]]` - put this above any function you want to make available in `R`
* `NumericVector` useful type implemented in `Rcpp`]

---
# Calling `C++` from `R`
.pull-left[
```{r eval=FALSE}
sourceCpp("mean.cpp")
```


```{r}
library(microbenchmark)
x <- runif(1e5)
microbenchmark(
  mean(x),
  meanC(x)
)

```
]


.pull-right[
* `sourceCpp` compiles the file
* More `Rcpp` syntax [here](https://dirk.eddelbuettel.com/code/rcpp/Rcpp-quickref.pdf)
]

---

class: middle

# Typing and modules

---

# Typing and methods
.pull-left-narrow[
- Scalar classes
* `int`
* `double`
* `bool`
* `String`
- vector classes
* `IntegerVector`
* `NumericVector`
* `LogicalVector`
* `CharacterVector`
]

.pull-right-wide[
- Vector classes have useful static methods (called with `::` on class) 

```{Rcpp eval=FALSE}
NumericVector v = NumericVector::create(1,2,3);
NumericVector::get_NA();
```

- Member functions or methods (called with `.` on object)
```{Rcpp eval=FALSE}
v.size();
v.push_back(4);
```
appends argument to the end of the `NumericVector v` 
]


---
# Typing and methods example

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector attribs() {
  NumericVector out = NumericVector::create(1, 2, 3);

  out.names() = CharacterVector::create("a", "b", "c");
  out.attr("my-attr") = "my-value";
  out.attr("class") = "my-class";

  return out;
}
```

---
# `SEXP` and `SEXPREC`

- All `R` objects are stored in a common datatype `SEXP` or S - expression.
 - This is a variant type with subvariants of the types you might expect (e.g. real, int, logical)
- Technically, a `SEXP` is a pointer to a structure of type `SEXPREC`

---
# Peeking at `R SEXP`

```{r eval=FALSE}
library(pryr)
#> Registered S3 method overwritten by 'pryr':
#>   method      from
#>   print.bytes Rcpp

sexp_type(10L)
#> [1] "INTSXP"
sexp_type("a")
#> [1] "STRSXP"
sexp_type(T)
#> [1] "LGLSXP"
sexp_type(list(a = 1))
#> [1] "VECSXP"
sexp_type(pairlist(a = 1))
#> [1] "LISTSXP"
#> 
```

.footnote[from http://adv-r.had.co.nz/C-interface.html]

---
# Casting using `as` and `wrap`
.pull-left[
```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

template <typename T>
T meanC(std::vector<T> x) {
  int n = x.size();
  float total = 0;
  
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}

// [[Rcpp::export]]
SEXP mean_wrap(SEXP input){
  std::vector<double> x = as<std::vector<double>>(input);
  float mean = meanC(x);
  
  return wrap(mean);
}
```
]
.pull-right[
```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;

  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}

```
]

---

---
# Including `Rcpp` in your `R` package

---

# `Rcpp` modules



---

# `Rcpp` sugar

---
