---
title:  "RCpp introduction"
subtitle: "FIMS Implementation Team Workshop"
author: "Christine Stawitz<br>NOAA Fisheries, OST<br>Seattle, WA<br> 2022-03-23"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "slides-style.css", "tufte.css"]
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---
layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-tile-view, echo=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()
```

---
# Sources
- "Advanced R"  by Hadley Wickham (http://adv-r.had.co.nz/Rcpp.html)
- "Rcpp: Seamless R and C++ integration" by Dirk Eddelbuetel and Romain Francois, Journal of Statistical Software (https://www.jstatsoft.org/article/view/v040i08)

---

# What is `RCpp` and why should I care?

.bluebox[## Definition
"The `RCpp` package provides a consistent API for seamlessly accessing, extending or modifying R objects at the C++ level" - Eddelbuttel and Francois.
]

.yellowbox[## API
a set of functions and procedures allowing the creation of applications that access the features or data of an operating system, application, or other service.]

* One of the most widely used `R` extensions (over 1000 packages)
* With very minimal knowledge of `C++` you can speed up function runs A LOT
- The most efficient `R` functions are written in `C++` and called from `R`


---
# Writing `C++` functions in `R`

* You can write `C++` functions inline in your R code using `cppFunction`
* You can also compile single lines of code directly using `evalCpp`

```{r}
library(Rcpp)
#> 
#> Attaching package: 'Rcpp'
#> The following object is masked from 'package:inline':
#> 
#>     registerPlugin
cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')
# add works like a regular R function
add
add(1, 2, 3)
evalCpp("std::numeric_limits<double>::max()")
```

---
# Calling `C++` from `R`

.pull-left[
```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;

  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}

```
]

.pull-right[
* `#include <Rcpp.h>` allows you to call `Rcpp` syntax in the file. You can omit `using namespace Rcpp` if you use `::` i.e. `Rcpp::NumericVector` 
* `// [[Rcpp::export]]` - put this above any function you want to make available in `R`
* `NumericVector, .size()` useful type, function implemented in `Rcpp`]

---
# Calling `C++` from `R`
.pull-left[
```{r eval=FALSE}
sourceCpp("mean.cpp")
```


```{r}
library(microbenchmark)
x <- runif(1e5)
microbenchmark(
  mean(x),
  meanC(x)
)

```
]


.pull-right[
* `sourceCpp` compiles the file
* More `Rcpp` syntax [here](https://dirk.eddelbuettel.com/code/rcpp/Rcpp-quickref.pdf)
]
---
# Compiling and running

* Compile your code using the `sourceCpp` `R` function.
```{r eval=FALSE}
sourceCpp("test.cpp")

```


---

class: middle

# Typing and modules

---

# Typing
.pull-left-wide[

```{Rcpp}
#include <Rcpp.h>

RcppExport SEXP convolve3cpp(SEXP a, SEXP b) {
Rcpp::NumericVector xa(a);
Rcpp::NumericVector xb(b);
int n_xa = xa.size(), n_xb = xb.size();
int nab = n_xa + n_xb - 1;
Rcpp::NumericVector xab(nab);
for (int i = 0; i < n_xa; i++)
for (int j = 0; j < n_xb; j++)
xab[i + j] += xa[i] * xb[j];
return xab;
}
```

]

.pull-right-narrow[
- `SEXP` This is an `R` type.
- 
]


---
# `SEXP` and `SEXPREC`

- All `R` objects are stored in a common datatype `SEXP` or S - expression.
 - This is a variant type with subvariants of the types you might expect (e.g. real, int, logical)
- Technically, a `SEXP` is a pointer to a structure of type `SEXPREC`
- All `C++` functions you write must take a `SEXP` as input and return an `SEXP`
---
# Peeking at `R SEXP`

```{r eval=FALSE}
library(pryr)
#> Registered S3 method overwritten by 'pryr':
#>   method      from
#>   print.bytes Rcpp

sexp_type(10L)
#> [1] "INTSXP"
sexp_type("a")
#> [1] "STRSXP"
sexp_type(T)
#> [1] "LGLSXP"
sexp_type(list(a = 1))
#> [1] "VECSXP"
sexp_type(pairlist(a = 1))
#> [1] "LISTSXP"
#> 
```

.footnote[from http://adv-r.had.co.nz/C-interface.html]


---

# `Rcpp` modules



---

class: middle

# A slide with just a title

---
